---
title: "vignette"
output:
  word_document: default
  pdf_document: default
vignette: "%\\VignetteIndexEntry{vignette} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hlacoloc)
```

## Introduction

HLA-coloc performs colocalization in two steps, as described in the manuscript. These are:

1. HLA gene causal signature: this uses SuSiE to assign a posterior inclusion probability (PIP) to each HLA allele.

2. HLA PIP Bayesian regression: this uses stanR to perform Bayesian regression of the PIPs in each genes separately.

## Required inputs

The following dataframes are necessary to run it (for each of the two phenotypes):

1. The HLA association summary statistics. This is a dataframe with the following columns:
  - `Name`: the name of the allele. These must be given in IMGT-HLA format, but can be of any field resolution though (as long as it's the same for both phenotypes).
  - `z`: the z-scores of the associations.
  - `beta`: the beta of the associations (if `z` is not provided).
  - `se`: the standard errors (if `z` is not provided).
  - `N`: the sample sizes.

2. The LD matrix (can be either in dataframe or matrix format). This is a square matrix of R coefficient for the HLA alleles above. Importantly, the order of the alleles needs to be the same in the summary statistics dataframe and in the LD matrix here.

## Example

### First, load the data:
The data is included in the package.
```{r}
data("ebna","ms","r_ebna","r_ms")
```


### Data snippet
Here's a snippet of the EBNA data:
```{r}
head(ebna)
r_ebna[1:5,1:5]
```

And here's a snippet of the multiple sclerosis (MS) data:
```{r}
head(ms)
r_ms[1:5,1:5]
```

### Now we run HLA-coloc on EBNA and MS:
This performs the 2 steps described above in one command.
```{r, results = "hide"}
coloc_res<-hla_coloc(pheno1=ebna,
                    pheno1R=r_ebna,
                    pheno1_name="EBNA",
                    pheno2=ms,
                    pheno2R=r_ms,
                    pheno2_name="Multiple sclerosis")
```

### Results:
Results include the colocalization table and a plot (optional).

The colocalization table includes 5 columns
  - `gene`: the HLA gene.
  - `susie_coloc_prob`: the probability that SuSiE selects at least one shared HLA allele between the two traits at the given gene.
  - `bayes_pd`: the probability that the SuSiE posterior inclusion probabilities (PIP) correlate for each genes.
  - `direction_of_correlation`: a check to make sure that the correlation between the PIPs is positive.
  - `hla_colocalization_probability`: the HLA colocalization probability (only valid if `direction_of_correlation` is correct). This is the product of `susie_coloc_prob` and `bayes_pd`, and is the final probability of HLA colocalization.

```{r, fig.width = 8.5, fig.height = 11}
coloc_res[["hla_colocalization"]]
```

The plot has two columns.
Left (a): the betas of the HLA allele summary statistics for the two traits.
Right (b): the PIPs obtained from SuSiE, and the final probabilities of HLA colocalization.

```{r, fig.width = 8.5, fig.height = 11}
coloc_res[["plot"]]
```


Lastly, if interested, users can retrieve the intermediate output from SuSiE, as a dataframe.
```{r, fig.width = 8.5, fig.height = 11}
head(coloc_res[["susie_intermediate_out"]])
```
