---
title: "vignette"
#output: rmarkdown::html_vignette
output:
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hlacoloc)
```

## Introduction

HLA-coloc performs colocalization in two steps, as described in the manuscript. These are:

1. HLA gene causal signature: this uses SuSiE to assign a posterior inclusion probability (PIP) to each HLA allele.

2. HLA PIP Bayesian regression: this uses stanR to perform Bayesian regression of the PIPs in each genes separately.

## Required inputs

The following dataframes are necessary to run it (for each of the two phenotypes):

1. The HLA association summary statistics. This is a dataframe with four columns:
  - `Name`: the name of the allele. These must be given in IMGT-HLA format, but can be of any field resolution though (as long as it's the same for both phenotypes).
  - `beta`: the beta of the associations.
  - `se`: the standard errors.
  - `N`: the sample sizes.

2. The LD matrix (can be either in dataframe or matrix format). This is a square matrix of R coefficient for the HLA alleles above. Importantly, the order of the alleles needs to be the same in the summary statistics dataframe and in the LD matrix here.

## Example

First, load the data:
```{r}
data("ebna","ms","r_ebna","r_ms")
```

Here's a snippet of the EBNA data:
```{r}
head(ebna)
r_ebna[1:5,1:5]
```

And here's a snippet of the MS data:
```{r}
head(ms)
r_ms[1:5,1:5]
```

Now we run coloc:
```{r, results = "hide"}
coloc_res<-hla_coloc(pheno1=ebna,
                    pheno1R=r_ebna,
                    pheno2=ms,
                    pheno2R=r_ms)
```

Results:
```{r, fig.width = 8.5, fig.height = 11}
coloc_res[["hla_colocalization"]]
coloc_res[["plot"]]
```
